\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-02-10 fluB.tex
Minor changes to text.

Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains and with non-human influenza A strains. 
Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are thought to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.

@@

We show that despite extensive reassortment, three of the eight segments, PB1, PB2 and HA, still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
This has occured through repeated reassortments and subsequent fixation of reassortant viruses within the influenza B population.

\begin{figure}[h]
 \centering		
	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
	\caption{\textbf{Schematic analysis of reassortment patterns.}
	A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages, in this case, the Victoria and Yamagata bifurcation that occurred in the early 1980s.
	B) We then transfer lineage labels from one tree to the same tips in another tree.
	Label switches thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation.
	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant.
	}
	\label{methodFig}
\end{figure}

\section*{Methods}
We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
The longest protein coding region of each segment was extracted and used for all further analyses.
%% GD - the following bit is meant to explain why we use M1 and NS1 alignments to make trees, but refer to them as MP and NS respectively in the text and figures, but I also realized that those labels have been used very inconsistently in the figures. I'll correct this when I re-make the figures to be slightly better resolution and also fix some minor typeface issues I've noticed. "M1" and "NS1" will only be used in the LD figure, since that is relevant if measuring amino acid LD, whereas whenever trees of M1 or NS1 are used we can be reasonably sure that the history of the MP and NS segment is accurately captured by M1 and NS1 regions.
We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
To date there has been little evidence of homologous recombination in influenza viruses \cite{chare2003,boni2008,han2010}.
The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.

Each strain was thus assigned 8 lineages depending on the combination of lineages it belongs to in all trees, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y.

@@

Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
Here, we modeled the substitution process using the HKY model of nucleotide substitution, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
We used a flexible Bayesian skyride demographic model \cite{minin2008}.
We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.

@@

We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
We do this by summing the branch lengths in each tree under 3 different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.

@@

We express the similarity of two trees by comparing the means of the distribution of TMRCA differences between pairs of tips in the two trees.
In comparing TMRCAs of trees A and B we calculate
\begin{equation}
\Delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
\end{equation}
where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.

These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.

Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
Approximate SPR distances were normalized using the procedure described above.

\subsection*{Linkage disequilibrium across the influenza B genome}
We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
For quantifying LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
LD was estimated only at loci where each nucleotide allele was represented by at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
We also calculated mean LD for all pairs of segments to quantify the overall association between them.
Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, as it would be impossible to put these measurements into a genome-wide perspective of linkage disequilibrium using just the data at hand.

@@

\subsection*{Phylogenetic analyses reveal maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} (see Figure \ref{genomeGrid}).
Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment.
Overall, we see many old bifurcations within the trees of all segments.
Many of these correspond to reassortment events involving the PB1-PB2-HA segment complexes of Vic and Yam lineages, highlighting the role reassortment plays in the long-term evolution of influenza B viruses.

@@

By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
Increasing mean TMRCA of Vic and Yam lineage pairs for all 5 segments (Figure \ref{betweenDiversity}) indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments, i.e.\ PB1, PB2 and HA traits after 1997 only `meet' at the root of PB1, PB2 and HA trees.
In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1 and PB2 segments on one hand and HA on the other.
These reductions in diversity represent small clades with PB1 and PB2 segments from one lineage and the HA segment from another, which go extinct by 1997.
In addition, the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.

@@

\begin{figure}[h]
	\centering	
	\includegraphics[width=0.65\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
	\caption{\textbf{Ratio of lineages in the dataset.}
The ratio of sequences that derive from the original Victoria clade (red) to sequences that drive from the Yamagata clade (blue) lineage sequences in each segment over time.
Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
PB1, PB2 and HA segments maintain separate Victoria and Yamagata lineages.}
	\label{lineageRatiosOverTime}
\end{figure}

We reconstructed reassortment events that were detected by using lineage labels.
Figure \ref{railroadPlot} focuses only on reassortments that have occured after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV) in Figure \ref{railroadPlot}.
Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-B2-HA.
Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to B/Iowa/03/2002-like isolates.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has been derived from Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored red and lineages that derive from the original Yamagata clade colored blue.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
Lineage extinction dates are not shown accurately.}
	\label{railroadPlot}
\end{figure}

\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}

We find that PB1, PB2 and HA segments, in addition to not reassorting across the Vic-Yam lineage boundary since 1997 (Figure \ref{betweenDiversity}), form co-reassorting segment complexes derived entirely of Vic or Yam lineage segments.
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).

@@

We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating a larger dataset with only PB1, PB2, HA and NA sequences.

@@

\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw reassortment rate.

@@

In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, see Supplementary information) between all segments.
We note that our SPR distance analysis might simply lack power.
SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
In addition, we use approximate SPR distances, rather than exact SPR distances.
However, we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (see Supplementary information).

Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
Most segment pairs show very low values for this statistic, indicating considerable differences between TMRCAs of the trees being compared.
A large number of these comparisons have TMRCA differences up to 10 times larger ($\Delta$TMRCA$\approx$0.1) than TMRCA differences that occur between replicate BEAST analyses of one alignment over the course of the MCMC chain.
PB1, PB2 and HA trees, on the other hand, exhibit normalized $\Delta$TMRCA values that are much higher.
This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs within either of these segments.
Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ there is isolation by distance.
In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).

@@

\subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
To estimate LD no other information, except for alleles at polymorphic amino acid sites in the dataset, is required.
Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
In Figure \ref{segmentLD} their signature is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
In some cases this value exceeds the estimated LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA).

@@

>>>>

Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to the ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains. 
Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are widely considered to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.

@@

We show that despite extensive reassortment, three of the eight segments, PB1, PB2 and HA, still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
This has occured through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.

\begin{figure}[h]
 \centering		
	\includegraphics[width=0.45\textwidth]{figures/TreeFigure2}
	\caption{\textbf{Schematic analysis of reassortment patterns.}
	A) We begin by assigning sequences falling on either side of a specified bifurcation within each segment tree to different lineages, in this case, the Victoria and Yamagata bifurcation that occurred in the early 1980s.
	B) We then transfer lineage labels from one tree to the same tips in another tree.
	Label switches thus indicate reassortment events that combine lineages falling on different sides of the Vic/Yam bifurcation.
	C) A reassortment graph depiction shows that tip number 6 is determined to be a reassortant based on B).
	}
	\label{methodFig}
\end{figure}

\section*{Methods}
We compiled a dataset of 452 complete influenza B genomes from GISAID \cite{GISAID} dating from 1984 to 2012 (accession numbers and laboratory acknowledgements can be found in Supplementary information).
The longest protein coding region of each segment was extracted and used for all further analyses.
We thus assume that homologous recombination has not taken place and that the evolutionary history of the whole segment can be inferred from the longest coding sequence in the segment.
To date there has been little evidence of homologous recombination in influenza viruses \cite{chare2003,boni2008,han2010}.
The segments of each strain were assigned to either Vic or Yam lineage by making maximum likelihood trees of each segment using PhyML \cite{guindon2003} and identifying whether the isolate was more closely related to B/Victoria/2/87 or B/Yamagata/16/88 sequences in that segment, with the exception of the NS segment (B/Victoria/2/87 was a reassortant and possessed a Yam lineage NS \cite{lindstrom1999}), where B/Czechoslovakia/69/1990 was considered as being representative of Victoria lineage.
Each strain was thus assigned 8 lineages depending on the combination of lineages from which their genomes were derived, for example all segments except for NS in strain B/Victoria/2/87 belong to Vic lineage and can thus be represented as V,V,V,V,V,V,V,Y.

@@

Temporally-calibrated phylogenies were recovered for each segment using the Markov chain Monte Carlo (MCMC) methods in the BEAST software package \cite{drummond2012}.
Here, we modeled the substitution process using the HKY model of nucleotide substitution \cite{hky1985}, with separate transition models for each of the 3 codon partitions, and additionally estimate realized synonymous and non-synonymous substitution counts \cite{obrien2009}.
We used a flexible Bayesian skyride demographic model \cite{minin2008}.
We accounted for incomplete sampling dates for 94 sequences (of which 93 only had year of isolation and 1 had only year and month of isolation) whereby tip date is estimated as a latent variable in the MCMC integration.

@@

We also calculated the total amount of evolutionary time spent by each segment with entirely Vic, entirely Yam or mixed lineage PB1, PB2 and HA segments.
We do this by summing the branch lengths in each tree under 3 (2 in the case of PB1, PB2 and HA trees) different lineage combinations of the PB1, PB2 and HA segments: PB1-PB2-HA derived entirely from Yamagata lineage, PB1-PB2-HA entirely derived from Victoria lineage and PB1-PB2-HA derived from a mixture of the two lineages.
This gives a measure of how successful, over long periods of time, each particular PB1-PB2-HA constellation has been.

@@

We express the similarity of two trees by comparing the means of the distribution of TMRCA differences between pairs of tips in the two trees.
In comparing TMRCAs of trees A and B we calculate
\begin{equation}
\Delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
\end{equation}
where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n |g(A_{ij}, B_{ij})|$ and $|g(A_{ij},B_{ij})|$ is the absolute difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A (which we refer to as A'), to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.
We had 3 replicate analyses of each segment and in order to calculate $f(A_i,A'_i)$ we used analyses numbered 1, 2 and 3 as A and analyses numbered 2, 3 and 1 as A', in that order.

Calculating $\Delta(A_i, B_i)$ for each MCMC state provides us with a posterior distribution of this parameter allowing  specific hypotheses regarding similarities between the trees of different segments to be tested.
Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.

Subtree prune and regraft (SPR) distances between phylogenetic trees are an approximate measure of the numbers of reassortment or recombination events.
Exact SPR distances are difficult to compute, as they depend on the SPR distance itself and are impractical to compute for posterior distributions of trees except for the most similar trees.
We calculated approximate SPR distances \cite{whidden2009,whidden2010,whidden2013} to quantify the numbers of reassortments that have taken place between all pairs of segments.
Approximate SPR distances were normalized using the procedure described above, where $f(A_i, A_i')$, $f(B_i, B_i')$ and $f(A_i, B_i)$ are approximate SPR distances between \textit{i}th posterior samples from trees A, B and indepent analyses thereof (A' and B').

\subsection*{Linkage disequilibrium across the influenza B genome}
We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
For quantifying LD we used the $\chi^{2}_{df}$ statistic \cite{zhao2005}, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
LD was estimated only at loci where each nucleotide or amino acid allele was represented by at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
We also calculated mean LD for all pairs of segments to quantify the overall association between them.
Linkage disequilibrium was not estimated for the larger PB1, PB2 and HA dataset, as it would be impossible to put these measurements into a genome-wide perspective of linkage disequilibrium using just the data at hand.

@@

\subsection*{Phylogenetic analyses reveal maintenance of diversity and co-assortment of PB1, PB2 and HA gene segments}
The differentiation into Vic and Yam lineages can be seen in all segments \cite{chen2008} (see Figure \ref{genomeGrid}).
Following the split of the two lineages, each segment can be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some segment lineages associated with mixed-lineage genomes have become fixed in the influenza B virus population, i.e.\ became the `trunk' of the phylogenetic tree of a segment.
Overall, we see many old bifurcations within the trees of all segments.
Many of these correspond to reassortment events involving the PB1, PB2 and HA segments of Vic and Yam lineages, highlighting the role reassortment plays in the long-term evolution of influenza B viruses.

@@

By measuring mean pairwise diversity between branches in each tree that were assigned either a Vic or Yam label in other segments, we look for reductions in between lineage diversity, which indicate that an inter-lineage reassortment event has taken place (Figure \ref{betweenDiversity}).
Because we are taking the mean of pairwise comparisons, this method gives a qualitative measure of panmixis between Victoria and Yamagata lineages in two trees.
We focus only on PB1, PB2 and HA lineage labels, since all other segments have fixed either the Vic or the Yam lineage.
Losses of diversity (represented by more recent mean pairwise TMRCAs between Vic and Yam labels) in Figure \ref{betweenDiversity} in all 8 segment trees indicate that every segment has reassorted with respect to the Victoria and Yamagata lineages of PB1, PB2 and HA segments.
However, we also see that PB1, PB2 and HA labels show reciprocal preservation of diversity after 1997.
This suggests that no reassortment events have taken place between Victoria and Yamagata lineages of PB1, PB2 and HA segments.
After 1997 in PB1, PB2 and HA trees PB1, PB2 and HA traits only `meet' at the root.
In a time period close to the initial split of Vic and Yam lineages (1986--1996) we see reduced diversity between Vic and Yam labels of PB1, PB2 and HA segments.
These reductions in diversity represent small clades with reassortant PB1-PB2-HA constellations, which go extinct by 1997.
We also observe that the assignment of Vic or Yam lineages of PB1, PB2 and HA segments to branches of other segment trees is very similar and often identical.
This results in PB1, PB2 and HA lineage labels switching between Vic and Yam simultaneously in all trees after 1997, suggesting co-reassortment of Vic and Yam lineages of PB1, PB2 and HA segments.

@@

\begin{figure}[h]
	\centering	
	\includegraphics[width=0.65\textwidth]	{figures/InfB_LineageRatiosOverTime.png}
	\caption{\textbf{Ratio of lineages in the dataset.}
The ratio of sequences that derive from the original Victoria clade (orange) to sequences that drive from the Yamagata clade (grey) lineage sequences in each segment over time.
Yamagata lineage PA, NP, NA and MP segments and Victoria lineage NS segment eventually become fixed in the influenza B population.
PB1, PB2 and HA segments maintain separate Victoria and Yamagata lineages.}
	\label{lineageRatiosOverTime}
\end{figure}

We reconstructed reassortment events that were detected by using lineage labels.
Figure \ref{railroadPlot} focuses only on inter-lineage reassortments that have occured after 1990.
We identify 5 major reassortant genome constellations (given in order PB1-PB2-PA-HA-NP-NA-MP-NS) circulating 1992 -- 2011: B/Alaska/12/1996-like (YYYYYYYV), B/Nanchang/2/1997-like (VVYVYVYV), B/Iowa/03/2002-like (VVYVYYYV), B/California/NHRC0001/2006-like (VVYVYYYV) and B/Brisbane/33/2008-like (VVYVYYYV) in Figure \ref{railroadPlot}.
Of these 5 constellations 4 (B/Nanchang/2/1997-like, B/Iowa/03/2002-like, B/California/NHRC0001/2006-like and B/Brisbane/33/2008-like) are derived from introgression of Yamagata lineage segments into Victoria lineage PB1-PB2-HA background, with only 1 (B/Alaska/12/1996-like) resulting from introgression of Victoria lineage NS segment into an entirely Yamagata lineage derived background.
All 5 inter-lineage reassortment events described here are marked by the preservation of either entirely Victoria or Yamagata lineage derived PB1-PB2-HA segments.
Figure \ref{railroadPlot} also shows that reassorting segments appear to evolve with a considerable degree of autonomy.
For example, the NP lineage that entered a largely Victoria lineage derived genome and gave rise to the B/Nanchang/2/1997-like isolates continued circulating until 2010, even though other segments it reassorted with in 1995 -- 1996 (PA and MP) went extinct following the reassortments that led to the rise of viruses with B/Iowa/03/2002-like genome constellations.
A more extreme example is the NS segment, which in B/Iowa/03/2002-like isolates (and all subsequent Vic PB1-PB2-HA isolates) has originally been derived from the Victoria lineage that had been associated with mostly Yam lineage derived B/Alaska/12/1996-like genomes for a number of years.

\begin{figure}[h]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/RailroadPlotDated.pdf}
	\caption{\textbf{Schematic plot of reconstructed reassortments between Victoria and Yamagata lineage segments of influenza B virus.}
Lineages that coassort in genomes are represented by 8 parallel lines, with lineages that derive from the original Victoria clade colored orange and lineages that derive from the original Yamagata clade colored grey.
Inter-lineage reassortment events are indicated by lines entering a different genome.
The angle of incoming lineages represents uncertainty in the timing of the event (mean date of the reassortant node and its parent node).
Lineage extinction dates are not shown accurately.}
	\label{railroadPlot}
\end{figure}

\subsection*{Recent PB1-PB2-HA constellations are entirely derived from Victoria or Yamagata lineage segments}
Figure \ref{stateTime} shows the sum of branch lengths which were labelled as having entirely Vic, entirely Yam or mixed-lineage PB1, PB2 and HA segments.
Due to lack of reassortment between Vic and Yam lineages of PB1, PB2 and HA (Figure \ref{betweenDiversity}) since 1997 all segments have spent significantly longer periods of evolutionary time with either entirely Vic-derived or entirely Yam-derived than with mixed-lineage PB1, PB2 and HA constellations (Figure \ref{stateTime}).

@@

We have identified 3 instances of mixed-lineage PB1-PB2-HA reassortants from the data with the following PB1-PB2-HA constellations: VVY (B/Bangkok/163/1990-like, 13 sequences isolated 1990 -- 5 Jan 1995), VYV (B/Nanchang/630/1994-like, 2 sequences isolated 1994 -- 1996) and VYY (B/New York/24/1993-like, 2 sequences isolated 8 Jan 1993 -- 1994).
An additional instance of reassortants with a YYV PB1-PB2-HA constellation (B/Waikato/6/2005-like, 16 sequences isolated 9 June -- 12 November in 2005) was discovered when investigating the larger dataset with only PB1, PB2 and HA sequences.

@@

\subsection*{PB1, PB2 and HA trees experience reassortment `distance' effects}
SPR (subtree prune and regraft) distances between perfectly resolved pairs of trees would be equal to the number of reassortments that have taken place between the two trees and is a measure of tree-to-tree similarity.
Figure \ref{SPRdistances} shows approximate SPR distances between all pairs of segment trees after normalization (see Methods).
If there are biases in the way segments reassort, so that some segments tend to co-reassort more often, we expect to see small-scale similarities between phylogenetic trees of those segments.
In our case we expect SPR distances, which are proportional to the number of reassortment events that have taken place between trees, to reflect the raw (\textit{i.e} both within and between lineages) reassortment rate.

@@

In Figure \ref{SPRdistances} the 95\% highest posterior density (HPD) intervals of normalized approximate SPR distances between pairs of segments encompass most means and occupy a relatively small range, suggesting there is no evidence of differences in the number of reassortments (or reassortment rate given as number of SPR moves per total time in the tree, see Supplementary information) between all segments.
We note, however, that our SPR distance analysis might simply lack power.
Analyses between independent analyses of the same segments yield distances that are comparable to distances between different segments (see Figures \ref{SPRdistancesTrees} and \ref{SPRdistancesReplicates}), suggesting that phylogenetic uncertainty is making a considerable contribution to our estimates of approximate SPR distances.
In addition, SPR distances themselves can only approximate (and underestimate) the actual numbers of reassortments.
Thus we caution against over-interpreting Figure \ref{SPRdistances}.
Although there might be concern about using approximate, rather than exact, SPR distances we do estimate exact SPR distances for a limited number of segment pairs - PB1, PB2 and HA - and find that after normalization exact and approximate SPR distances are not significantly different (see supplementary Figures \ref{NormSPR_PB1-PB2_difference}--\ref{NormSPR_PB2-HA_correlation}).

Although we do not find evidence of differences in numbers of raw reassortments between all pairs of trees, we find support for a reassortment `distance' effect.
Figure \ref{deltaTMRCA} shows normalized mean $\Delta$TMRCA values for all pairs of trees.
Most segment pairs show very low values for this statistic, indicating considerable differences between TMRCAs of the trees being compared.
A large number of these comparisons have TMRCA differences up to 10 times larger ($\Delta$TMRCA$\approx$0.1) than TMRCA differences that occur between replicate BEAST analyses of one alignment over the course of the MCMC chain.
PB1, PB2 and HA trees, on the other hand, exhibit normalized $\Delta$TMRCA values that are much higher.
This is an indicator of TMRCA differences between trees of PB1, PB2 and HA segments which, though noisy, are occassionally very similar to uncertainty in tip-to-tip TMRCAs between replicate analyses of these segments.
Overall, it suggests that PB1, PB2 and HA lineages tend to not reassort amongst themselves unless both reassorting segments have similar TMRCAs, i.e.\ there is isolation by distance.
In addition, we see evidence of a similar reassortment `distance' effect in MP and NA trees in Figure \ref{deltaTMRCA}.
Previously, associations between MP and NA segments have been suspected in influenza A viruses (Samantha Lycett, personal communication).

@@

\subsection*{Linkage disequilibrium analysis confirms phylogenetic findings}
Linkage disequilibrium (LD) is a measure of association between alleles at different polymorphic loci within a population.
To estimate LD no other information, except for alleles at polymorphic amino acid sites in the dataset, is required.
Although not an independent validation of association between PB1, PB2 and HA segments, linkage disequilibrium analyses show that Victoria and Yamagata lineages of these segments have accumulated lineage-specific amino acid substitutions.
In Figure \ref{segmentLD} their signature is greater mean $\chi^{2}_{df}$ between PB1, PB2 and HA amino acid sites than any other pair of proteins.
In some cases this value exceeds the estimated mean LD of amino acid sites within the same protein (\textit{e.g.} PB1 and HA have higher mean LD than mean LD when PB1 is compared against itself).

@@

\begin{figure}
\centering  
\includegraphics[width=0.65\textwidth]  {supp_figures/InfB_robustCounting.png}
\caption{\textbf{Synonymous, non-synonymous and nucleotide substitution rates in segments under different PB1-PB2-HA complexes.}}
\label{robustCounting}
\end{figure}
